04-analisis

### 2024-07-03 04-analisis.qmd
correccion analisis jc

@@

>>>>

```
---
author: "Equipo Edumer"
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/apa6.csl"
---

@@

@fig-bivariate shows a series of graphs depicting the association between the variables of market justice preferences - in education, health, and pensions - and the variables of meritocratic perception at school (effort and talent) and in society (effort, talent, and deservingness) (see conceptual diagram in @fig-hypotheses). In general we observe a positive association between meritocracy and market justice preferences for the societal questions, but this is not as straightforward for the school items, particularly for the one referred to effort.

This section presents the results of the multilevel models. We begin by showing the results of the cumulative link mixed models in @tbl-ordinal-reg for the ordinal dependent variables of justice in health, pensions, and education separately in order to compare the effect of the predictors. Then, we present the results of the linear mixed-effects models for the average index of preference for market justice in @tbl-lineal-reg[^1].

For the health item, the intraclass correlation (ICC) obtained for the null model (not shown) indicates that 5% of the total variance of this item is associated with specific characteristics of the schools, while it is 3% for the pensions item and 4% for the education item (ICC based on [@hox_multilevel_2010, p.15]). For the average index of market justice preferences, the ICC reaches 4%, indicating that overall, only a small percentage of the total variance in students' market justice preferences is associated with school characteristics, thus limiting the possibility of finding effects at this second level of analysis.

[^1]: The table shows the effects of individual and contextual-level independent variables related to our hypotheses; however, the models also incorporate the control variables for each of the three dependent variables separately and for the average index (complete models are available in the [Appendix](#appendix)).

Additionally, Model 1 examines family status through parents' educational level and the number of books at home. For health, pensions, and education, a negative but statistically non-significant association with parents' education is observed. However, having more than 25 books at home shows a negative and statistically significant effect on market justice in health and pensions, suggesting less justification for differentiated access compared to those with fewer books.

Model 2 in @tbl-ordinal-reg incorporates school-level variables. The results show that high socioeconomic level schools show greater preferences for market justice in health, pensions, and education compared to low socioeconomic level schools, with statistically significant results (*p* \< 0.05). Moreover, and contrary to our hypothesis, schools with higher academic performance show negative and statistically significant effects across the three items.

@tbl-lineal-reg presents the results of linear mixed-effects models for the market justice preference index. Model 1 shows that higher perceived societal meritocracy, through effort, talent, and deservingness, is positively and significantly associated with market justice preferences. School meritocracy perceptions yield mixed effects: perceived talent increases market justice preference, while perceived effort decreases it, both statistically significant (*p* \< 0.001). Model 2 includes family status, finding no statistically significant effects from parents' education or the number of books at home. Regarding school characteristics, Models 3 and 4 introduce socioeconomic level and academic performance. The results of Model 3 suggest a statistically significant relationship between school socioeconomic levels and market justice preferences, specifically those schools with high socioeconomic levels ($\beta$ = 0.37) compared to low socioeconomic levels. However, Model 4 indicates that schools with medium ($\beta$ = -0.11) and high ($\beta$ = -0.26) academic performance present lower market justice preferences compared to low-performing schools, with these effects being statistically significant when controlling for other variables.

@tbl-interact displays the interaction terms related to our hypotheses. The results suggest that students' family background moderates the relationship between their meritocratic perceptions and their justification of market allocation in social services. Specifically, the relationship between social effort and market justice preferences is less positive for students whose parents have university or postgraduate education. Lower-status students (measured by parental education) prefer more market justice when they strongly adhere to meritocratic perceptions of deservingness. A similar moderating effect is seen with family cultural capital: the relationship between effort and market justice preferences is less positive for students with more than 25 books at home.
